### 2011-10-14 doc/manuscript.tex
An interesting vim extension for saving sessions.

In addition to these biological categories, we also found several aminoacyl tRNA
synthetases (GO Term translation GO:0006412) in association with recombining
gene transfer.  In the parsimony-based approach the term translation was top
ranked (Table S\ref{tab:go-events-recombining}).  They had been a focus of gene
transfer in bacterial species \citep{Woese2000}.  Some of the functional
categories were unexpected; functional categories concerning ribosomes were
found to be associated with recombining gene transfer.  Because genes related
with ribosomes were considered to be housekeeping genes, gene transfer in which
would be rare, it was unexpected that ribosome related genes were found to be
associated with recombining gene transfer.  Because it was a bizarre result, we
investigated the association by considering different recombinant edge types. We
noted that two different classes of recombinant edge types: one that changes
species tree topologies, and another that does not. Because recombinant edges
that did not change species tree topology might be difficult to infer reliably,
considering those recombinant edges in computing recombinantion intensity might
lead to dubious results.  Ribosome related functional categories were found to
be top ranked if we took those recombinant edges that did not change species
tree topology (Tables S\ref{tab:functional-notopology} and
S\ref{tab:functional-all}).  In Table \ref{tab:functional} ribosome related
functional categories were not top ranked.  Therefore, high recombination
intensity in ribosome related genes were likely to be attributed to
topology-not-changing recombinant edges. We should be cautious of recombining
transfer of genes related with ribosome related functions.

The parsimony-based approach also showed
enrichment of the recombining transfers in many of the same categories such as
translation, ribosome, and rRNA binding when using all types of recombinant
edges in computing recombination intensity (see Tables
S\ref{tab:functional-all} and S\ref{tab:go-events-recombining}).  
Because the
parsimony-based method did not take species tree branch lengths into account, 
the agreement between the two different methods might be likely.

For horizontal gene transfer we found enrichment in very different categories
related to transposition, integration, and recombination (Table
\ref{tab:go-events}).  Gene duplication and loss events were enriched for
categories similar to those of horizontal gene transfer's enrichment (Tables
S\ref{tab:go-events-duplication} and S\ref{tab:go-events-loss}).
Gene duplication and horizontal transfers showed enrichment in the same
categories perhaps because both mechanisms increase gene copy number.  
Gene losses showed enrichment for transposition and integration, but less
significantly (P-value of $<10^{-6}$), and showed enrichment for several
additional categories such as transporter activity and transcription initiation.

Comparing two types of recombinant edges, one from SDE to SPY and another from
SPY to SDE transport related categories were more associated for the direction
from SPY to SDE
(Table S\ref{tab:functional-sde-spy} and S\ref{tab:functional-spy-sde}).

\texttt{More interpretations about Mawgli analysis of function categories?
Why transposases?}

@@

We investigated closely related bacterial species to better understand gene
transfer processes between SPY and SDE. Two complementary methods were employed
to study recombining and horizontal gene transfers. By discriminating the two
processes in analyzing the genomes, we showed that the pathogenicity of SDE and
SPY was more attributed to horizontal gene transfer process than recombining
gene transfer.  Because recombination in bacteria seemed to occur at somewhat
random positions along the genome, the homologous parts transferred might be
unlikely to be enriched with virulence factors to human hosts.  Recombining gene
transfer might have rarely contributed to influx of virulence genes from SPY to
the opportunistic human pathogen of SDE as we showed a few cases of virulence
gene that might have been transferred via recombining gene transfer.

We also showed that the net amount of recombining gene flows from SPY to SDE was
relatively greater than that of the reverse direction in the core genome
sequences.  If there were unbalanced recombining gene transfer between the two
species, what were the barriers \citep{Thomas2005} to the gene transfer in the
bacteria? Although barriers to bacterial recombination would include geographic
isolation, genetic isolation could prevent recombination between coinhabiting
bacteria.  \citet{Kalia2001} discussed restriction-modification system (RMS) as
a possible mechanism of the unbalanced gene flow.  Restriction-modification
system is a biological self-defense mechanism of bacteria, which is diverse in
prokaryote world.  Biological roles of RMS might be to maintain and control
species identity \citep{Jeltsch2003}.  Recently, \citet{Budroni2011a} studied
that \textit{Neisseria meningitidis} phylogenetic clades were associated with
RMS, claiming that RMS modulated homologous recombination in the bacteria of
\textit{N.\ meningitidis}.  Most of \textit{S.\ pyogenes} genomes contain
restriction-modification type I system as well as type II system. The two
genomes of SDE contained only type II system. More tight self-defense system of
SPY may have made the species less capable of aquiring foreign genetic material
from SDE, and accordingly relatively unbalanced gene flow between SPY and SDE
may have been associated with different restriction-modification systems in the
two species \citep{Kalia2001}.

We used relatively small number of genomes to detect recombining gene transfer
that had only been possible using many samples for a few genes. We revisited the
problem of gene transfer in other previous studies.  We reexamined the seven
genes that \citet{Kalia2001} had studied for net directionality of gene transfer
between SPY and SDE.  Internal fragments of the six housekeeping genes, encoding
putative glucose kinase (\textit{gki}), glutamine transport protein
(\textit{gtr}), glutamate racemase (\textit{murI}), mismatch repair enzyme
(\textit{mutS}), transketolase (\textit{recP}), and xanthine
phosphoribosyltransferase (\textit{xpt}) were located in the core genome that we
investigated in this study.  Genes \textit{mutS}, \textit{recP}, and
\textit{xpt} had fragments of gene transfer with moderate posterior probability
of recombination. The other genes did not have noticeable recombination. Except
for \textit{mutS} where \citet{Kalia2001} did not find recombination, the result
was consistent with that of \citet{Kalia2001}.  We also revisited recombination
in gene \textit{parC} in SPY1 genome studied by \citet{Pinho2010}, where they
studied fluoroquinolone resistance in SDE and reported recombination in the
gene. We observed moderate recombination in the gene.

We used two heterogeneous methods: one model-based and another parsimony-based
for studying recombining and horizontal gene transfer.  The core genomes
consisted of sequence alignments with rare gaps. With this inherent filtering of
the input data preparation the model-based method did not accommodate gene
dupliation and loss in its core genome analysis. On the contrary, the
parsimony-based method used all of the genomes not just the core genomes.
Considering nucleotides as the level of gene duplication, loss, and transfer
towards a complete model for recombining and horizontal gene transfer could be
difficult because genome-scale application of those models based on DNA or
protein sequences might not be feasible.  Combining the two approaches for
studying the mode of gene transfer would be desirable to study more complicated
evolutionary processes including recombining gene transfer, gene duplication,
loss, and horizontal gene transfer.

\texttt{Limitation of ClonalOrigin: inability of inferring recombinant edges}

\texttt{Need of a new model with asymmetry of gene transfer}

\texttt{Need of more SDE genomes}

@@

\subsection{Simulation Study}

We conducted experiments on simulation data to check the accuracy of estimates
obtained by \texttt{ClonalOrigin} in a scenario similar to one from the real 
data set.  We used a
simulation feature of \texttt{ClonalOrigin} to simulate the process of mutation
and recombination across the five species 
(see section \textbf{Supplementary Information} for detail).
While populaiton mutation and
recombination parameters were accurately estimated, the recombinant tract length
was over-estimated  (Table S\ref{tab:sim-three-population-parameters}).  This
was not surprising as recombinant tract lengths were known to be difficult to
estimate from relatively short alignment blocks \citep{Didelot2010}. 
We also evaluated the accuracy of the statistics estimated in the second stage
of \texttt{ClonalOrigin}. The number of recombinant edges for each pair of
recombinant tree branches was correlated with the true value in general while
there were slight over-estimation (Figure S\ref{fig:h1}).  However, this
comparison between inferrence from simulated data and that of real data set
suggested that the directional difference between SPY and SDE was due to neither
lack of power of \texttt{ClonalOrigin} nor hidden bias thereof.  Recombination
intensities that were measured on genes also showed rough correlation (Pearson's
correlation coefficient of 0.63) of the estimated and true values (Figure
S\ref{fig:ri1}).  Estimated values appeared to be driven somewhat by the prior
implied by the population parameter setting, explaining why lower values were
over-estimated and higher values were under-estimated.  Nonetheless, the
correlation indicated that the measure of recombination intensity used in our
functional gene category served as an adequate indicator of actual intensity of
recombination events along genes.

@@

\subsection{Local gene trees and recombination intensity}

@@

The recombination intensity allowed us to find genomic regions enriched for
recombination events in general, or regions enriched for recombination events of
certain types (e.g., between SPY and SDE).  We noted that the recombination
intensity was not normalized to the interval of $[0,1]$. Rather than giving
probabilistic meanings to the recombination intensity we interpreted it as an
average number of recombinant edge types per site.  The measure appeared to act
as an effective indicator of recombination intensity (see subsection of
\textbf{Simulation Study}).  Our simulation study indicated that edges of
certain types were better indicators of recombination events than others.
For instance, recombinant edges which resulted in a recombinant tree with the
same topology as the species tree appeared to be more prone to spurious
sampling.  Such edges were ones where the donor was the same branch, sister
branch, or parent branch of the recipient branch.  Therefore, in some cases we
elected to remove such edges from consideration when measuring recombination
intensity.  
% The tree topologies changes would be correlated with recombination
% intensity.  The tree topologies would show more effectively regions where
% recombination occurs along the genome (see subsection of \textbf{UCSC Genome
% Browser}).

\subsection{Population mutation and recombination rates were accurately
estimated}

% FIXME: Newick tree in text in a file.
% FIXME: Publish the result and the data set with the code.
We describe the detailed procedure of simulations for checking if the three
population parameters were accurately recovered from the first stage of
\texttt{ClonalOrigin} MCMC.  Ten recombinant trees were sampled given the three
parameters and the species tree  set to the values inferred from the five
bacterial genomes.  Each of the recombinant trees was used to simulate the
Jukes-Cantor model for DNA sequence evolution to create 274 blocks of
alignments.  The block lengths configuration was the same as that of the blocks
used in the analysis of the five genomes.  The mean of the ten estimates of the
recombinant tract length was significantly higher than the true value of (919 bp
versus 744 bp) whereas the values of the mutation and recombination rates were
accurately estimated (Table S\ref{tab:sim-three-population-parameters}).
Recombinant tract lengths were known to be difficult to estimate from relatively
short alignment blocks \citep{Didelot2010}.

\subsection{Estimating numbers of recombinant edges were accurate}

% FIXME 1: IG: RESULTS FROM THIS POINT ON REQUIRE SOME REFINEMENT. SC: Let's see.
% FIXME 2: ``TRUE" TO BE EXPECTED VALUE
% RATHER THAN AVERAGE OVER THE 10 REC.  TREES. SC: OKAY. LET'S FIND THE PRIOR
% EXPECTED NUMBER OF EVENTS. THIS MUST BE GIVEN SOMEWHERE}.  
% FIXME 3: CONSIDER SHOWING DETAILED COMPARISON OF ESTIMATED VALUES TO ONES 
% EXPECTED FROM THE ACTUAL REC. TREES. THESE DON'T LOOK AS NICE, BUT THEY MIGHT 
% BE ABLE TO DEMONSTRATE OUR ABILITY TO DETECT DEPARTURE FROM EXPECTATION. SC:
% THIS MAY BE ABOUT CHANGING AVERAGE TO PRIOR EXPECTED NUMBER OF EVENTS.
% FIXME 4: WE SHOULD PROBABLY CONSIDER ALTERNATIVE WAYS OF USING THESE RESULTS TO 
% ARGUE THAT OUR DIRECTIONAL SIGNAL IS REAL. SC: I MIGHT CONSIDER SIMULATIONS AND
% ARTIFICIALLY DOUBLE OR TRIPLE THE RECOMBINATION EVENTS OF ONE OF TYPES.
Because of the two-stage Bayesian method we first sampled ten recombinant trees
given the mutation rate, recombination rate, and tract length, and then created
ten independent simulated data sets with the Jukes-Cantor model for DNA sequence
evolution using each of the recombinant trees. This consisted of 100 replicates
of data sets.  Figure S\ref{fig:h1} shows the observed count of each recombinant
edge type against the expected count; the estimated numbers appeared to be
slightly over-estimated, but in correlation with the true numbers.

\subsection{Unbalanced gene flow model was adequate}

We checked if the model with unbalanced gene flow was adequate using posterior
predictive check. We used 100 recombinant trees from the posterior distribution
of the recombiant tree given the real data of the 274 alignment blocks in order
to simulate the ClonalOrigin model to create a new data set of 100 274-block
alignments.  Using the simulated data sets of 100 274-block alignments we
infer recombinant trees to draw the distribution of logarithm base 2 ratio of
observed recombinant edge counts to the prior expected number of recombinant
edges. If the model with unbalanced gene flow was adequate, the logarithm base 2
ratio from the real data set should center around the distribution. 
\texttt{What is the result of this simulation?}

\subsection{Simulation for recombination intensity}

We tested if our measure of recombination intensity provided a reliable estimate
of the true intensity. For this purpose, we artificially annotated our synthetic
blocks to genes according to the real annotations along the real bacterial
alignments. For each gene in each of the simulated 100 data sets, we plotted the
average estimated recombination intensity along that gene against the true
average intensity, i.e., the average number of different recombinant edge types
implied by the recombinant tree used in simulation 
(Pearson's correlation coefficient of 0.63 and Figure S\ref{fig:ri1}).

@@

%=============================================================================
% Table S. ClonalOrigin - SDE-to-SPY
\begin{table}
\caption{\label{tab:functional-sde-spy}Functional categories asscoiated with 
recombination intensity based on SDE-to-SPY recombinant edge types.}
\noindent \begin{centering}
\begin{tabular}{cccl}
\hline 
P-Value & Count & GO term & Description\tabularnewline
\hline 
1.854e-05 & 38 & GO:0009401 & phosphoenolpyruvate-dependent sugar phosphotransferase system\tabularnewline
2.703e-05 & 81 & GO:0006412 & translation\tabularnewline
6.482e-05 & 33 & GO:0008982 & protein-N(PI)-phosphohistidine-sugar phosphotransferase activity\tabularnewline
9.777e-05 & 105 & GO:0003723 & RNA binding\tabularnewline
0.00012793 & 40 & GO:0030529 & ribonucleoprotein complex\tabularnewline
0.00025176 & 26 & GO:0019843 & rRNA binding\tabularnewline
0.00031236 & 72 & GO:0005975 & carbohydrate metabolic process\tabularnewline
0.00044756 & 317 & GO:0005737 & cytoplasm\tabularnewline
0.00059675 & 45 & GO:0008643 & carbohydrate transport\tabularnewline
0.00066874 & 39 & GO:0003735 & structural constituent of ribosome\tabularnewline
0.00135304 & 10 & GO:0006098 & pentose-phosphate shunt\tabularnewline
\hline 
\end{tabular}
\par\end{centering}
\end{table}
\clearpage{}
%=============================================================================

%=============================================================================
% Table S. ClonalOrigin - SPY-to-SDE
\begin{table}
\caption{\label{tab:functional-spy-sde}Functional categories asscoiated with 
recombination intensity based on SPY-to-SDE recombinant edge types.}
\noindent \begin{centering}
\begin{tabular}{cccl}
\hline 
P-Value & Count & GO term & Description\tabularnewline
\hline 
1.106e-06 & 45 & GO:0008643 & carbohydrate transport\tabularnewline
1.511e-06 & 33 & GO:0008982 & protein-N(PI)-phosphohistidine-sugar phosphotransferase activity\tabularnewline
1.595e-06 & 38 & GO:0009401 & phosphoenolpyruvate-dependent sugar phosphotransferase system\tabularnewline
1.443e-05 & 187 & GO:0006810 & transport\tabularnewline
0.0001106 & 371 & GO:0016740 & transferase activity\tabularnewline
0.0003596 & 72 & GO:0005975 & carbohydrate metabolic process\tabularnewline
0.0004023 & 13 & GO:0006814 & sodium ion transport\tabularnewline
0.000823 & 317 & GO:0005737 & cytoplasm\tabularnewline
0.001142 & 12 & GO:0031402 & sodium ion binding\tabularnewline
\hline 
\end{tabular}
\par\end{centering}
\end{table}
\clearpage{}
%=============================================================================

%=============================================================================
% Table S. Mowgli - Duplication 
\begin{table}
\caption{\label{tab:go-events-duplication}Functional categories enriched in 
gene duplication events that are inferred using the parsimony-based approach.}
\noindent \begin{centering}
\begin{tabular}{cccl}
\hline 
P-value & Count & GO-term & Description \\
\hline 
4e-81 &  67 & GO:0006313 & transposition, DNA-mediated\\
5e-80 &  59 & GO:0004803 & transposase activity\\
6e-70 &  23 & GO:0032196 & transposition\\
2e-58 &  52 & GO:0015074 & DNA integration\\
8e-25 &  86 & GO:0006310 & DNA recombination\\
1e-20 &  36 & GO:0016987 & sigma factor activity\\
1e-20 &  36 & GO:0006352 & transcription initiation\\
2e-19 & 197 & GO:0003676 & nucleic acid binding\\
2e-10 & 527 & GO:0003677 & DNA binding\\
4e-06 & 142 & GO:0043565 & sequence-specific DNA binding\\
\hline 
\end{tabular}
\par\end{centering}
\end{table}
\clearpage{}
%=============================================================================

>>>>

In addition to these biological categories, we also found several aminoacyl tRNA
synthetases (GO Term translation GO:0006412) in association with recombining
gene transfer.  In the parsimony-based approach the term translation was top
ranked (Table S\ref{tab:go-events-recombining}).  They had been a focus of gene
transfer in bacterial species \citep{Woese2000}.  Some of the functional
categories were unexpected; functional categories concerning ribosomes were
found to be associated with recombining gene transfer.  Because genes related
with ribosomes were considered to be housekeeping genes, gene transfer in which
would be rare, it was unexpected that ribosome related genes were found to be
associated with recombining gene transfer.  Because it was a bizarre result, we
investigated the association by considering different recombinant edge types. We
noted that two different classes of recombinant edge types: one that changes
species tree topologies, and another that does not. Because recombinant edges
that did not change species tree topology might be difficult to infer reliably,
considering those recombinant edges in computing recombinantion intensity might
lead to dubious results.  Ribosome related functional categories were found to
be top ranked if we took those recombinant edges that did not change species
tree topology (Tables S\ref{tab:functional-notopology} and
S\ref{tab:functional-all}).  In Table \ref{tab:functional} ribosome related
functional categories were not top ranked.  Therefore, high recombination
intensity in ribosome related genes were likely to be attributed to
topology-not-changing recombinant edges. We would be cautious of recombining
transfer of genes related with ribosome related functions.

\subsection{Simulations}

1. Topology-not-changing recombinant edge was unreliably estimated compared with
topology-changing recombinant edge.
2. Balanced case
3. Unbalanced case

We conducted experiments on simulation data to check the accuracy of estimates
obtained by the model-based approach in a scenario similar to one from the real 
data set.

While populaiton mutation and
recombination parameters were accurately estimated, the recombinant tract length
was over-estimated  (Table S\ref{tab:sim-three-population-parameters}).  
Recombinant tract lengths were reported to be difficult to estimate from relatively
short alignment blocks \citep{Didelot2010}.

Figure S\ref{fig:h1} shows the observed count of each recombinant
edge type against the expected count; the estimated numbers appeared to be
slightly over-estimated, but in correlation with the true numbers.

We also evaluated the accuracy of the statistics estimated in the second stage
of \texttt{ClonalOrigin}.

The number of recombinant edges for each pair of
recombinant tree branches was correlated with the true value in general while
there were slight over-estimation (Figure \ref{fig:h3}).

However, this
comparison between inferrence from simulated data and that of real data set
suggested that the directional difference between SPY and SDE was due to neither
lack of power of \texttt{ClonalOrigin} nor hidden bias thereof.

Recombination
intensities that were measured on genes also showed rough correlation (Pearson's
correlation coefficient of 0.63) of the estimated and true values (Figure
S\ref{fig:ri1}).

Estimated values appeared to be driven somewhat by the prior
implied by the population parameter setting, explaining why lower values were
over-estimated and higher values were under-estimated.

Nonetheless, the
correlation indicated that the measure of recombination intensity used in our
functional gene category served as an adequate indicator of actual intensity of
recombination events along genes.

@@

Molecular evolution of bacterial genomes is multifaceted. Studies of bacterial
evolution often focused on one side of many evolutionary processes.  While
understanding one aspect of bacterial evolutionary processes we could be
ignorant of the other features.  Horizontal gene transfer process was often
considered in macro-evolutionary scale using a great deal of bacterial species.
As closely related species or groups of bacterial individuals are becoming a
focus of evolutionary study, recombining gene transfer process would surface as
another facet of evolutionary forces.  Because the two gene transfer processes
would act on bacterial genomes simultaneously, ignoring one of the two processes
in studying evolution of bacteria would lead to a narrowed point of view to the
underlying evolutionary forces. Our study of SPY and SDE genomes undertook the
work of discriminating recombining and non-recombining horizontal gene transfer
processes because the two species were very closely related.  We showed that the
two gene transfer processes acted on different biological functions. Recombining
gene transfer process in the analysis of SPY and SDE genomes was related with
genes encoding membrane-bound molecules and aminoacyl tRNA synthetases. On the
contrary, horizontal gene transfer process was found to be associated with
transposases, which would allow horizontal gene transfer of their neighboring
genes.  Because SPY was a human-specific pathogen, and SDE was an opportunistic
human pathogen, we were interested in association of pathogenicity with the two
gene transfer processes. We showed that virulence genes were more attributed to
horizontal gene transfer process than to recombining gene transfer process.
With the current data of only SPY and SDE genomes we could not conclude that the
association of virulence genes with horizontal gene transfer not recombining
gene transfer was applicable to other species.  However, without the
discrimination of the two processes it would have been difficult to find the
disparate association of biological functions or virulence genes with the
different gene transfer processes.

Another contribution of our study to bacterial gene transfer research is to
elucidate the net directionality of gene flow from SPY to SDE in the core
genomes. Although the issue of net directionality has been tackled, our study
was the first attempt of resolving the issue in genome-wide scale. Yet, it still
remains to be seen whether this net directionality of gene flow from SPY to SDE
happened in the pan genome. A model-based method with more refined models would
be necessary for the reserach direction because the parsimony-based method we
employed simply could not show any net directionality between SPY and SDE in the
pan genome. Although we are unsure whether much larger number of genomes could
be useful in finding net directionality, we consider that more SDE genomes would
be beneficial to future studies.

Were there unbalanced recombining gene transfer between the two species, what
were the barriers \citep{Thomas2005} to cause the unbalanced gene transfer in
the species?  \citet{Kalia2001} discussed restriction-modification system (RMS)
as a possible mechanism of the unbalanced gene flow.  Restriction-modification
system is a biological self-defense mechanism of bacteria, which is diverse in
prokaryote world.  Biological roles of RMS might be to maintain and control
species identity \citep{Jeltsch2003}.  Recently, \citet{Budroni2011a} studied
that \textit{Neisseria meningitidis} phylogenetic clades were associated with
RMS, claiming that RMS modulated homologous recombination in the bacteria of
\textit{N.\ meningitidis}.  Most of \textit{S.\ pyogenes} genomes contained
restriction-modification type I system as well as type II system. The two
genomes of SDE contained only type II system. More tight self-defense system of
SPY may have made the species less capable of aquiring foreign genetic material
from SDE, and accordingly relatively unbalanced gene flow between SPY and SDE
may have been associated with different restriction-modification systems in the
two species \citep{Kalia2001}.

We reexamined the seven genes that \citet{Kalia2001} had studied for net
directionality of gene transfer between SPY and SDE.  The internal fragments of
the six housekeeping genes (\textit{gki}, \textit{gtr}, \textit{murI},
\textit{mutS}, \textit{recP}, \textit{xpt}) were located in the core genome that
we investigated in this study.  Except for \textit{mutS} where \citet{Kalia2001}
did not find recombination, our results were consistent with that of
\citet{Kalia2001}.  \citet{Pinho2010} reported recombination in the gene
\textit{parC}, in which we also observed moderate recombination.

It remains to be seen whether the inability of detecting net directionality
between SPY and SDE in the pan genomes was because of lack of power of the
parsimony-based approach that we employed. Considering gene duplication, loss,
and horizontal gene transfer in a model-based approach would be desirable in the
coming ages of abundant bacterial genomes. The model might consider genome
alignments where each taxon does not necessarily have a unique gene; some taxa
might have no gene due to gene loss, and other taxa multiple genes due to gene
gain, duplication, or horizontal gene transfer. With these types of models will
we be able to address more interesting evolutionary questions in bacteria. This
need of more refined models would be also true even for the study of net
directionality between species in the core genome.  In other words, one could
envision a model where a population two-taxa (i.e., SPY and SD groups) tree is
superimposed on the species tree shown in Figure S\ref{fig:clonalorigin}A. A
splitting event of the population tree could happen somewhere between the root
of the species tree and the internal node of SPY. Recombinant edges that connect
branches within SPY or SD clades would be more frequent than those that connect
branches between SPY and SDE clades.  With the reference to the recombination
within a clade one parameter can dictate the degree of gene from from SPY to
SDE, and another for the reverse direction.  If the two parameters could be
incorporated in the model, then they might be used to compare the degrees of
gene flow between the two species in a more statistically sound way.  We hope
that we will see thie refinement of the model to better illustrate bacterial
evolution in forthcoming bacterial genomic era.

@@

\subsection{Simulation Study}
We performed three sets of simulations.  In the first simulation, we wished to
show that \texttt{ClonalOrigin} could infer population mutation rate, tract
length, and recombination rate with reasonable accuracy. In the second
simulation, we wished to check whether recombinant edges were recovered from
simulated data with recombinant trees generated from the prior. The third
simulation was conducted using simulated data with recombinant trees sampled the
posterior distribution.  We used a simulation feature of \texttt{ClonalOrigin}
to simulate the process of mutation and recombination across the five species
(see section \textbf{Supplementary Information} for detail).

@@

\subsection{Summarization of posterior samples for local gene trees and recombination intensity}

@@

The recombination intensity allowed us to find genomic regions enriched for
recombination events in general, or regions enriched for recombination events of
certain types (e.g., between SPY and SDE).  We noted that the recombination
intensity was not normalized to the interval of $[0,1]$. Rather than giving
probabilistic meanings to the recombination intensity we interpreted it as an
average number of recombinant edge types per site.  The measure appeared to act
as an effective indicator of recombination intensity (see subsection of
\textbf{Simulation Study}).  Our simulation study indicated that edges of
certain types were better indicators of recombination events than others.  For
instance, recombinant edges which resulted in a recombinant tree with the same
topology as the species tree appeared to be more prone to spurious sampling.
Such edges were ones where the donor was the same branch, sister branch, or
parent branch of the recipient branch.  Therefore, in some cases we elected to
remove such edges from consideration when measuring recombination intensity.  
% The tree topologies would show more effectively regions where
% recombination occurs along the genome (see subsection of \textbf{UCSC Genome
% Browser}).

\subsection{Simulation 1 - three population parameters}

We describe a procedure of simulations for checking if the three
population parameters were accurately recovered using the model-based approach.
Ten recombinant trees were sampled given the three parameters (Table
S\ref{tab:three}) and the species tree (Figure S\ref{fig:tree5}) set to the
values inferred from the five bacterial genomes.  Each of the recombinant trees
was used to simulate the Jukes-Cantor model for DNA sequence evolution to create
274 blocks of alignments.  The block lengths configuration was the same as that
of the blocks used in the analysis of the five genomes.

% s15
% FIXME: Newick tree in text in a file.
% FIXME: Publish the result and the data set with the code.

\subsection{Simulation 2 - numbers of recombinant edges from prior}

We first sampled ten recombinant trees given the mutation rate, recombination
rate, and tract length (Table S\ref{tab:three}) and the species tree (Figure
S\ref{fig:tree5}), and then created ten independent simulated data sets with
the Jukes-Cantor model for DNA sequence evolution using each of the recombinant
trees. This consisted of 100 replicates of data sets.  The recombinant trees
were created from the prior distribution and recombinant edges were added to the
species tree with no particular directions of recombining gene transfers between
species tree branches.  We computed the ratio of the mean number of recombinant
edges for a pair of donor and recipient species tree branches over the simulated
10 recombinant trees against the median of recombinant edges estimated from the
10 simulated data sets, and then took the logarithm of the ratio to base 2. We
also computed 5\% and 95\% quantiles to find the 90\% quantile range for the log
base 2 ratio.  We expect that the log base 2 ratio would near zero. We also
wished to see if how the recombinant edge count estimates from the real data set
deviate from the prior. Instead of using the number of recombinant edges from
the ten recombinant trees, we used the numbers of recombinant edges that were
estimated from the real data.

We tested if our measure of recombination intensity provided a reliable estimate
of the true intensity. We artificially annotated our synthetic blocks to genes
according to the real annotations from SPY1 genome.  For each gene in each of
the simulated 100 data sets, we plotted the average estimated recombination
intensity along that gene against the average number of different recombinant
edge types from the ten simulated recombinant tree.

% s16

% s16
% FIXME 1: IG: RESULTS FROM THIS POINT ON REQUIRE SOME REFINEMENT. SC: Let's see.
% FIXME 2: ``TRUE" TO BE EXPECTED VALUE
% RATHER THAN AVERAGE OVER THE 10 REC.  TREES. SC: OKAY. LET'S FIND THE PRIOR
% EXPECTED NUMBER OF EVENTS. THIS MUST BE GIVEN SOMEWHERE}.  
% FIXME 3: CONSIDER SHOWING DETAILED COMPARISON OF ESTIMATED VALUES TO ONES 
% EXPECTED FROM THE ACTUAL REC. TREES. THESE DON'T LOOK AS NICE, BUT THEY MIGHT 
% BE ABLE TO DEMONSTRATE OUR ABILITY TO DETECT DEPARTURE FROM EXPECTATION. SC:
% THIS MAY BE ABOUT CHANGING AVERAGE TO PRIOR EXPECTED NUMBER OF EVENTS.
% FIXME 4: WE SHOULD PROBABLY CONSIDER ALTERNATIVE WAYS OF USING THESE RESULTS TO 
% ARGUE THAT OUR DIRECTIONAL SIGNAL IS REAL. SC: I MIGHT CONSIDER SIMULATIONS AND
% ARTIFICIALLY DOUBLE OR TRIPLE THE RECOMBINATION EVENTS OF ONE OF TYPES.

\subsection{Simulation 3 - number of recombinant edges from posterior}

We used 100 recombinant trees sampled from the posterior distribution
of the recombiant tree instead of using simulated recombinant trees.
We created one simulated data set for each of the 100 recombinant trees with
the Jukes-Cantor model for DNA sequence evolution 
using the mutation rate estimate (Table S\ref{tab:three}).
This consisted of 100 replicates of data sets.  
We follow the same computation of the ratio as in the simulation 2 with 
the numbers of recombinant edges estimated from the real data
as numerators.

% s17

@@

%=============================================================================
% Table S. ClonalOrigin - SDE-to-SPY
%\begin{table}
%\caption{\label{tab:functional-sde-spy}Functional categories asscoiated with 
%recombination intensity based on SDE-to-SPY recombinant edge types.}
%\noindent \begin{centering}
%\begin{tabular}{cccl}
%\hline 
%P-Value & Count & GO term & Description\tabularnewline
%\hline 
%1.854e-05 & 38 & GO:0009401 & phosphoenolpyruvate-dependent sugar phosphotransferase system\tabularnewline
%2.703e-05 & 81 & GO:0006412 & translation\tabularnewline
%6.482e-05 & 33 & GO:0008982 & protein-N(PI)-phosphohistidine-sugar phosphotransferase activity\tabularnewline
%9.777e-05 & 105 & GO:0003723 & RNA binding\tabularnewline
%0.00012793 & 40 & GO:0030529 & ribonucleoprotein complex\tabularnewline
%0.00025176 & 26 & GO:0019843 & rRNA binding\tabularnewline
%0.00031236 & 72 & GO:0005975 & carbohydrate metabolic process\tabularnewline
%0.00044756 & 317 & GO:0005737 & cytoplasm\tabularnewline
%0.00059675 & 45 & GO:0008643 & carbohydrate transport\tabularnewline
%0.00066874 & 39 & GO:0003735 & structural constituent of ribosome\tabularnewline
%0.00135304 & 10 & GO:0006098 & pentose-phosphate shunt\tabularnewline
%\hline 
%\end{tabular}
%\par\end{centering}
%\end{table}
%\clearpage{}
%=============================================================================

%=============================================================================
% Table S. ClonalOrigin - SPY-to-SDE
%\begin{table}
%\caption{\label{tab:functional-spy-sde}Functional categories asscoiated with 
%recombination intensity based on SPY-to-SDE recombinant edge types.}
%\noindent \begin{centering}
%\begin{tabular}{cccl}
%\hline 
%P-Value & Count & GO term & Description\tabularnewline
%\hline 
%1.106e-06 & 45 & GO:0008643 & carbohydrate transport\tabularnewline
%1.511e-06 & 33 & GO:0008982 & protein-N(PI)-phosphohistidine-sugar phosphotransferase activity\tabularnewline
%1.595e-06 & 38 & GO:0009401 & phosphoenolpyruvate-dependent sugar phosphotransferase system\tabularnewline
%1.443e-05 & 187 & GO:0006810 & transport\tabularnewline
%0.0001106 & 371 & GO:0016740 & transferase activity\tabularnewline
%0.0003596 & 72 & GO:0005975 & carbohydrate metabolic process\tabularnewline
%0.0004023 & 13 & GO:0006814 & sodium ion transport\tabularnewline
%0.000823 & 317 & GO:0005737 & cytoplasm\tabularnewline
%0.001142 & 12 & GO:0031402 & sodium ion binding\tabularnewline
%\hline 
%\end{tabular}
%\par\end{centering}
%\end{table}
%\clearpage{}
%=============================================================================

%=============================================================================
% Table S. Mowgli - Duplication 
%\begin{table}
%\caption{\label{tab:go-events-duplication}Functional categories enriched in 
%gene duplication events that are inferred using the parsimony-based approach.}
%\noindent \begin{centering}
%\begin{tabular}{cccl}
%\hline 
%P-value & Count & GO-term & Description \\
%\hline 
%4e-81 &  67 & GO:0006313 & transposition, DNA-mediated\\
%5e-80 &  59 & GO:0004803 & transposase activity\\
%6e-70 &  23 & GO:0032196 & transposition\\
%2e-58 &  52 & GO:0015074 & DNA integration\\
%8e-25 &  86 & GO:0006310 & DNA recombination\\
%1e-20 &  36 & GO:0016987 & sigma factor activity\\
%1e-20 &  36 & GO:0006352 & transcription initiation\\
%2e-19 & 197 & GO:0003676 & nucleic acid binding\\
%2e-10 & 527 & GO:0003677 & DNA binding\\
%4e-06 & 142 & GO:0043565 & sequence-specific DNA binding\\
%\hline 
%\end{tabular}
%\par\end{centering}
%\end{table}
%\clearpage{}
%=============================================================================

%=============================================================================
% Table S. Mowgli - Loss
%\begin{table}
%\caption{\label{tab:go-events-loss}Functional categories enriched in 
%gene loss events that are inferred using the parsimony-based approach.}
%\noindent \begin{centering}
%\begin{tabular}{cccl}
%\hline 
%P-value & Count & GO-term & Description \\
%\hline 
%1e-06 &  67 & GO:0006313 & transposition, DNA-mediated\\
%2e-05 &  59 & GO:0004803 & transposase activity\\
%5e-05 &  52 & GO:0015074 & DNA integration\\
%5e-04 &  23 & GO:0032196 & transposition\\
%5e-04 &  10 & GO:0046873 & metal ion transmembrane transporter activity\\
%7e-04 &  36 & GO:0016987 & sigma factor activity\\
%7e-04 &  36 & GO:0006352 & transcription initiation\\
%1e-03 &  31 & GO:0006814 & sodium ion transport\\
%3e-03 &  86 & GO:0006310 & DNA recombination\\
%6e-03 & 197 & GO:0003676 & nucleic acid binding\\
%\hline 
%\end{tabular}
%\par\end{centering}
%\end{table}
%\clearpage{}
%=============================================================================
